---
title: "ILEC data analysis"
author: "Michiel Luteijn"
date: '2025-12-16'
output:
  html_document:
    theme: cosmo
    toc: yes
    toc_float: yes
    code_folding: hide
---

# Background  {.tabset}

## Objectives
**Big Pitch submission** <br>
Health and care (H&C) actuaries are well positioned to benefit from recent advances in data science, as machine learning (ML) techniques have become increasingly transparent and accessible. <br>

Once seen as opaque, explainable ML now enables actuaries to improve predictive power whilst maintaining interpretability. Explainable ML can enhance traditional methods, whilst retaining the ability to infuse actuarial judgement.  <br>

We will demonstrate that H&C actuaries can have the best of both worlds by combining transparency, expert knowledge, and modern analytical innovation.

**IFoA webpage** <br>
TITLE: Breaking down the black box: why actuaries should embrace machine learning now. <br>
Description: Achieving the best of both worlds with explainable machine learning [H&C: Data Techniques in H&C WP].

## Decisions made
**Study design** <br>
- It was decided in Dec to build incidence models (i.e. from scratch with lives exposure as the offset) <br>
- Dec 19th - Provisionally decided to have the last 2 calendar years (2018-2019) as the holdout

**Study inclusion criteria** <br>
- Dec 19th - decided TERM only <br>
- Dec 19th - to take out fields with no exposure (none of these have claims) <br>
- Dec 19th - decided the cut-off for issue year should be 1980 onwards <br>
- Dec 19th - decided the all observation years (2012-2019) will be included <br>
- Dec 19th - decided the range for issueAge should be 18-80 <br>
- Dec 19th - decided the range for Attained_Age should be 18-90 <br>
- Dec 19th - No exclusions made based on combinations of SOA_Guar_Lvl_TP and duration (for now), or Face_Amount_Band <br>
- Jan 9th - decided the cap duration at 20 due to concerns about data quality at (very) high durations and low volumes of data <br>
- Jan 13th - decided the exclude unknown smoking status


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r package install, results=FALSE, message=FALSE, warning=FALSE, include = F}
packages <- c("data.table", "dplyr", "ggplot2", "magrittr", "purrr", "knitr", "stringr", "plyr", "htmlTable", "lme4", "statmod", "glmnet", "splines", "MASS", "rms", "caret", "foreach", "parallel", "readr", "arrow", "doSNOW", "here", "htmltools", "plotly", "xgboost", "SHAPforxgboost")

for (pkg in packages){
  if(!pkg %in% rownames(installed.packages())){
    install.packages(pkg)
  } 
}
library("purrr")
quietly(lapply(packages, library, character.only = T)) %>% suppressWarnings()

rm(packages, pkg)
quietly(gc())
```

```{r, include=FALSE, echo=FALSE}
set.seed(1234)
```

```{r, include=FALSE, echo=FALSE}
library(plotly)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(fig.showtext = TRUE)
```

```{r, include=FALSE, echo=FALSE}
# Source https://stackoverflow.com/questions/60685631/using-ggplotly-and-dt-from-a-for-loop-in-rmarkdown
# Init Step to make sure that the dependencies are loaded
htmltools::tagList(DT::datatable(cars))
htmltools::tagList(ggplotly(ggplot()))
```

```{r Register parallel backend, include=FALSE}
if(!exists("cl")){
  cl <- parallel::makeCluster(8, type = "SOCK") # Creating a SOCK cluster
  doSNOW::registerDoSNOW(cl)
}
```

```{r functions, include=FALSE}
helpers <- list.files(path = here::here("R/Helpers"), pattern = "\\.R$", full.names = TRUE)
lapply(helpers, source)
```

```{r Data load, include=FALSE}
fpath = "Data/ILEC_2012_19 - 20240429.parquet"

if(!file.exists(here::here(fpath))){
  source("R/read_data_and_save_parquet.R")
  } else {
    ilec <- arrow::read_parquet(here::here(fpath))
    }
```

# Data filtering {.tabset}

## Filtering
- AtoEs based on SOA-supplied ExpDth_VBT2015wMI_Cnt that came with the dataset. <br>

```{r Data filters}
ilec <- log_detailed_summary(ilec, quote(!Policies_Exposed == 0), reason = "Taking out zero exposure fields")
ilec <- log_detailed_summary(ilec, quote(Issue_Year >= 1980), reason = "Removed IssueYear pre-1980")
ilec <- log_detailed_summary(ilec, quote(Attained_Age >= 18), reason = "Removed Attained_Age below 18")
ilec <- log_detailed_summary(ilec, quote(Attained_Age <= 90), reason = "Removed Attained_Age above 90")
ilec <- log_detailed_summary(ilec, quote(Issue_Age >= 18), reason = "Removed Issue_Age below 18")
ilec <- log_detailed_summary(ilec, quote(Issue_Age <= 80), reason = "Removed Issue_Age above 80")
ilec <- log_detailed_summary(ilec, quote(Duration <= 20), reason = "Removed duration above 20")
ilec <- log_detailed_summary(ilec, quote(Smoker_Status != "U"), reason = "Removed unknown smoking status")

DT::datatable(attributes(ilec)$filter_summary)
```

## Remaining
- AtoEs based on SOA-supplied ExpDth_VBT2015wMI_Cnt that came with the dataset. <br>

```{r Data remaining}
summary <- 
  ilec[, list(
    Deaths = sum(Death_Count),
    Claim_Amounts_Bn = (sum(Death_Claim_Amount)/1000000000) %>% round(4),
    Exposure_years_Mln = (sum(Policies_Exposed)/1000000) %>% round(4),
    Avg_claim = (sum(Death_Claim_Amount)/sum(Death_Count)) %>% round(-3),
    AtoE_lives = (sum(Death_Count)/sum(ExpDth_VBT2015wMI_Cnt)) %>% round(4) * 100
  )]

DT::datatable(summary)
```


```{r aggregate to speed up analytics}
group_cols <- c("Observation_Year", "Sex", "Smoker_Status", "Issue_Age", "Duration", "Face_Amount_Band", "Issue_Year", "Attained_Age")
measure_cols <- c("Death_Count", "Policies_Exposed", "ExpDth_VBT2015wMI_Cnt")

ilec <- ilec[, lapply(.SD, sum, na.rm = TRUE), by = group_cols, .SDcols = measure_cols]
```

```{r covert chars to factors}
char_cols <- names(ilec)[vapply(ilec, is.character, logical(1))]
if (length(char_cols)) {
  ilec[, (char_cols) := lapply(.SD, as.factor), .SDcols = char_cols]
}

cols_to_order <- c("Face_Amount_Band")
for (j in intersect(cols_to_order, names(ilec))) {
  x  <- ilec[[j]]
  lv <- order_band_levels(x)
  f  <- factor(x, levels = lv, ordered = TRUE)
  set(ilec, j = j, value = f)
}
```

```{r test train split}
years_sorted <- sort(unique(ilec$Observation_Year))
test_years <- tail(years_sorted, 2)
is_test <- ilec$Observation_Year %in% test_years
dt_train <- ilec[!is_test, , drop = FALSE]
dt_test <- ilec[ is_test, , drop = FALSE]
```

```{r feature engineering}
age_coefs <- attr(poly(dt_train$Attained_Age, 3), "coefs")
entry_age_coefs <- attr(poly(dt_train$Issue_Age, 3), "coefs")
obs_year_coefs <- attr(poly(dt_train$Observation_Year, 3), "coefs")

dt_train <- gen_features(dt_train, age_coefs, entry_age_coefs, obs_year_coefs)
dt_test <- gen_features(dt_test, age_coefs, entry_age_coefs, obs_year_coefs)
```

# Basic GLM {.tabset}
- First build age only. The age only model will be the offset for models downstream to allow for better interpretation of regression coefficients. 

```{r GLM}
f <- list()
f$age_fit <- stats::as.formula(
  Death_Count ~ 
    (agepoly1 + agepoly2 + agepoly3 + agespline2 + agespline3 + agespline4)
)
glm_age_fit <- glm(as.formula(f$age_fit), dt_train, family = "poisson", offset = log(Policies_Exposed))
```

## Age model
```{r }
DT::datatable(
  glm_rr_table(glm_age_fit),
  rownames = FALSE,
  extensions = "Buttons",
  options = glm_dt_options()
) %>% glm_dt_format()
```

```{r add pred}
dt_train[, c("age_pred") := predict(glm_age_fit, dt_train, newoffset = dt_train$Policies_Exposed) %>% exp]
dt_test[, c("age_pred") := predict(glm_age_fit, dt_test, newoffset = dt_test$Policies_Exposed) %>% exp]
```

```{r GLM basic}
f$basic <- stats::as.formula(
  Death_Count ~ (Sex * Smoker_Status + Face_Amount_Band + issue_agepoly1 + issue_agepoly2 + Duration + invDur + dur_sq + dur_cube + obs_yearpoly1 + obs_yearpoly2)
  )
glm_basic <- glm(as.formula(f$basic), dt_train, family = "poisson", offset = log(age_pred))
```

## Stacked basic model
```{r }
DT::datatable(
  glm_rr_table(glm_basic),
  rownames = FALSE,
  extensions = "Buttons",
  options = glm_dt_options()
) %>% glm_dt_format()
```

```{r add age model pred}
dt_train[, c("basic_pred") := predict(glm_basic, dt_train, newoffset = dt_train$age_pred) %>% exp]
dt_test[, c("basic_pred") := predict(glm_basic, dt_test, newoffset = dt_test$age_pred) %>% exp]
```

# Visuals of GLM fit on test data {.tabset}
- Visuals on the test data (calendar years 2018-2019)

## Age model {.tabset}
- Note AtoEs are low since experience is lighter in 2018-2019, than in prior observation years which is not accounted for in the age model. <br>

```{r }
ae_plots  <- setNames(vector("list", length(group_cols)), group_cols)
actual_col <- "Death_Count"
pred_col   <- "age_pred"

ae_plots <- setNames(
  lapply(group_cols, function(gc) {
    summ <- ae_summary_by(dt_test, gc)
    ae_plot(summ, gc, top_n = 50)
  }),
  group_cols
)
```

``` {r, results = 'asis', echo = F}
for(i in 1:length(ae_plots)){
  cat(paste0("### ", names(ae_plots)[i]," \n"))
  print(htmltools::tagList(plotly::ggplotly(ae_plots[[i]])))
}
```

## Basic GLM {.tabset}

```{r }
ae_plots  <- setNames(vector("list", length(group_cols)), group_cols)
pred_col   <- "basic_pred"

ae_plots <- setNames(
  lapply(group_cols, function(gc) {
    summ <- ae_summary_by(dt_test, gc)
    ae_plot(summ, gc, top_n = 50)
  }),
  group_cols
)
```

``` {r, results = 'asis', echo = F}
for(i in 1:length(ae_plots)){
  cat(paste0("### ", names(ae_plots)[i]," \n"))
  print(htmltools::tagList(plotly::ggplotly(ae_plots[[i]])))
}
```

# Build XGBoost to probe the GLM fit for deficiencies {.tabset}
- For the XGBoost, 2017 is used as test data, whilst 2012-2016 are train data <br>

## Model Fitting
```{r XGBoost}
feature_cols <- names(dt_train)[1:8]

# Train/validation split (same as your year-based split)
dt_tr <- dt_train[Observation_Year != 2017]
dt_va <- dt_train[Observation_Year == 2017]

# Targets and offsets
y_tr <- dt_tr[["Death_Count"]]
y_va <- dt_va[["Death_Count"]]
offset_tr <- log(dt_tr[["basic_pred"]])
offset_va <- log(dt_va[["basic_pred"]])

# Build a formula for the selected features
f$XGBoost <- as.formula(paste("~", paste(feature_cols, collapse = " + ")))

# One-hot encode
X_tr <- onehot_encode(dt_tr[, 1:8], f$XGBoost)
X_va <- onehot_encode(dt_va[, 1:8], f$XGBoost)

# DMatrices
dtrain <- xgb.DMatrix(data = X_tr, label = y_tr)
setinfo(dtrain, "base_margin", offset_tr)

dvalid <- xgb.DMatrix(data = X_va, label = y_va)
setinfo(dvalid, "base_margin", offset_va)

params <- list(
  objective = "count:poisson",
  eval_metric = "poisson-nloglik",
  eta = 0.1,
  max_depth = 2,
  subsample = 0.8,
  colsample_bytree = 0.8,
  min_child_weight = 100
  )

evals_list <- list(train = dtrain, validation = dvalid)

xgb_model <- xgb.train(
  params = params,
  data = dtrain,
  nrounds = 2000,
  evals = evals_list,
  early_stopping_rounds = 20,
  print_every_n = 10,
  verbose = 0
  )
```

## SHAP Feature Importance
```{r SHAP_importance}
shap_vals <- shap.values(xgb_model, X_va)

shap_importance <- data.frame(
  Feature = names(shap_vals$mean_shap_score),
  MeanAbsSHAP = shap_vals$mean_shap_score %>% round(4)
) %>% data.table

shap_importance <- shap_importance[order(-shap_importance$MeanAbsSHAP), ]
DT::datatable(shap_importance)
```

## SHAP Beeswarm Plot
```{r shap_summary, fig.width=10, fig.height=8}
shap_long <- SHAPforxgboost::shap.prep(
  shap_contrib = shap_vals$shap_score,
  X_train = X_va
)

SHAPforxgboost::shap.plot.summary(shap_long)
```

## SHAP Dependence Plots{.tabset}
```{r SHAP dependence, results='asis', echo = F, message = F, warning = F}
sub_idx <- sample(unique(shap_long$ID), size = 2000)
toPlot <- shap_importance[Feature != "(Intercept)"]$Feature
N_features <- pmin(length(toPlot), 10)

for(feature in toPlot[1:N_features]){
  cat(paste0("### ", feature," \n\n"))
  
  plot <- shap.plot.dependence(
    shap_long[ID %in% sub_idx],
    X = X_va[sub_idx,],
    x = feature,
    color_feature = feature_cols[2]
  )
  
  plot <- plot + 
    scale_y_log10(
      labels = function(x) round(exp(x), 4),
      name = "exp(SHAP value), A/E equivalent"
    )
  feature_values <- X_va[sub_idx, feature, with = FALSE] %>% unlist
  if(length(unique(feature_values)) == 2 && all(feature_values %in% c(0, 1, TRUE, FALSE))) {
    plot <- plot + 
      scale_x_continuous(
        breaks = c(0, 1),
        labels = c("FALSE", "TRUE")
      )
  }
  
  plotly_obj <- ggplotly(plot)
  print(htmltools::tagList(plotly_obj))
  cat('\n\n')
}
```

## SHAP Interaction Analysis
```{r shap_interactions, fig.width=12, fig.height=10}

```

## Interaction Heatmap
```{r SHAP dependence interactions, results='asis', echo = F, message = F, warning = F}
# Create interaction matrix (approximation)
#if (length(top_features_interact) >= 2) {
#  # Calculate pairwise SHAP interaction strengths
#  interaction_matrix <- matrix(0, 
 #                               nrow = length(top_features_interact), 
#                                ncol = length(top_features_interact))
#  rownames(interaction_matrix) <- top_features_interact
#  colnames(interaction_matrix) <- top_features_interact
  
#  for (i in 1:length(top_features_interact)) {
#    for (j in 1:length(top_features_interact)) {
#      if (i != j) {
#        # Correlation of SHAP values as proxy for interaction
#        interaction_matrix[i, j] <- abs(cor(
#          shap_values$shap_score[, top_features_interact[i]],
#          shap_values$shap_score[, top_features_interact[j]]
#        ))
#      }
#    }
#  }
  
  # Plot heatmap
#  library(reshape2)
#  melted_matrix <- melt(interaction_matrix)
  
#  ggplot(melted_matrix, aes(x = Var1, y = Var2, fill = value)) +
#    geom_tile() +
#    geom_text(aes(label = round(value, 2)), size = 3) +
#    scale_fill_gradient2(low = "white", high = "red", mid = "yellow", 
#                         midpoint = 0.5, limit = c(0, 1)) +
#    theme_minimal() +
#    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
#    labs(
#      title = "SHAP Value Correlation Matrix (Interaction Proxy)",
#      subtitle = "Identifying potential interaction terms missing from GLM_basic",
#      x = "", y = "", fill = "Abs Correlation"
#    )
#}
```

## Model Comparison
```{r lgbm_comparison}

```

## Key Insights
```{r insights}

```

# Evaluate model performance
``` {r, results = 'asis', echo = F}
models <- list(
  GLM_age   = list(
    train_pred = dt_train$age_pred, 
    test_pred = dt_test$age_pred
  ),
  GLM_basic = list(
    train_pred = dt_train$basic_pred, 
    test_pred = dt_test$basic_pred
  )
)

comparison <- imap_dfr(models, ~ data.table::data.table(
  Model             = .y,
  Poisson_dev_train = poisson_dev(dt_train$Death_Count, .x$train_pred),
  Poisson_dev_test  = poisson_dev(dt_test$Death_Count, .x$test_pred)
))

DT::datatable(comparison) |>
  DT::formatRound(columns = c("Poisson_dev_train", "Poisson_dev_test"), digits = 4)
```
