---
title: "ILEC data analysis"
author: "Michiel Luteijn"
date: '2025-12-16'
output:
  html_document:
    theme: cosmo
    toc: yes
    toc_float: yes
    code_folding: hide
---

# Background  {.tabset}

## Objectives
**Big Pitch submission** <br>
Health and care (H&C) actuaries are well positioned to benefit from recent advances in data science, as machine learning (ML) techniques have become increasingly transparent and accessible. <br>

Once seen as opaque, explainable ML now enables actuaries to improve predictive power whilst maintaining interpretability. Explainable ML can enhance traditional methods, whilst retaining the ability to infuse actuarial judgement.  <br>

We will demonstrate that H&C actuaries can have the best of both worlds by combining transparency, expert knowledge, and modern analytical innovation.

**IFoA webpage** <br>
TITLE: Breaking down the black box: why actuaries should embrace machine learning now. <br>
Description: Achieving the best of both worlds with explainable machine learning [H&C: Data Techniques in H&C WP].

## Decisions made
**Study design** <br>
- It was decided in Dec to build incidence models (i.e. from scratch with lives exposure as the offset) <br>
- Dec 19th - Provisionally decided to have the last 2 calendar years (2018-2019) as the holdout

**Study inclusion criteria** <br>
- Dec 19th - decided TERM only <br>
- Dec 19th - to take out fields with no exposure (none of these have claims) <br>
- Dec 19th - decided the cut-off for issue year should be 1980 onwards <br>
- Dec 19th - decided the all observation years (2012-2019) will be included <br>
- Dec 19th - decided the range for issueAge should be 18-80 <br>
- Dec 19th - decided the range for Attained_Age should be 18-90 <br>
- Dec 19th - No exclusions made based on combinations of SOA_Guar_Lvl_TP and duration (for now), or Face_Amount_Band <br>
- Jan 9th - decided the cap duration at 20 due to concerns about data quality at (very) high durations and low volumes of data <br>
- Jan 13th - decided the exclude unknown smoking status


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r package install, results=FALSE, message=FALSE, warning=FALSE, include = F}
packages <- c("data.table", "dplyr", "ggplot2", "magrittr", "purrr", "knitr", "stringr", "plyr", "htmlTable", "lme4", "statmod", "glmnet", "splines", "MASS", "rms", "caret", "foreach", "parallel", "readr", "arrow", "doSNOW", "here", "htmltools", "plotly", "xgboost", "SHAPforxgboost")

for (pkg in packages){
  if(!pkg %in% rownames(installed.packages())){
    install.packages(pkg)
  } 
}
library("purrr")
quietly(lapply(packages, library, character.only = T)) %>% suppressWarnings()

rm(packages, pkg)
quietly(gc())
```

```{r, include=FALSE, echo=FALSE}
set.seed(1234)
```

```{r flags, include=FALSE, echo=FALSE}
test_calendaryear <- FALSE # Sets test data be calendar years 2018-2019 (TRUE), or 20% random sample (FALSE)
data_proportion <- FALSE # Set to desired data sample fraction, otherwise FALSE (or 1)
depths <- 1:3 # Tree depths for IBLM XGBoost booster
```

```{r, include=FALSE, echo=FALSE}
library(plotly)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(fig.showtext = TRUE)
```

```{r, include=FALSE, echo=FALSE}
# Source https://stackoverflow.com/questions/60685631/using-ggplotly-and-dt-from-a-for-loop-in-rmarkdown
# Init Step to make sure that the dependencies are loaded
htmltools::tagList(DT::datatable(cars))
htmltools::tagList(ggplotly(ggplot()))
```

```{r Register parallel backend, include=FALSE}
if(!exists("cl")){
  cl <- parallel::makeCluster(8, type = "SOCK") # Creating a SOCK cluster
  doSNOW::registerDoSNOW(cl)
}
```

```{r functions, include=FALSE}
helpers <- list.files(path = here::here("R/Helpers"), pattern = "\\.R$", full.names = TRUE)
lapply(helpers, source)
```

```{r Data load, include=FALSE}
fpath = "Data/ILEC_2012_19 - 20240429.parquet"

if(!file.exists(here::here(fpath))){
  source("R/read_data_and_save_parquet.R")
  } else {
    ilec <- arrow::read_parquet(here::here(fpath))
  }
```

```{r partial run, include=FALSE}
if(!isFALSE(data_proportion)){
  ilec <- ilec[sample.int(.N, .N * data_proportion)]
  }
```

# Data filtering {.tabset}

## Filtering
- AtoEs based on SOA-supplied ExpDth_VBT2015wMI_Cnt that came with the dataset. <br>

```{r Data filters}
ilec <- log_detailed_summary(ilec, quote(!Policies_Exposed == 0), reason = "Taking out zero exposure fields")
ilec <- log_detailed_summary(ilec, quote(Issue_Year >= 1980), reason = "Removed IssueYear pre-1980")
ilec <- log_detailed_summary(ilec, quote(Attained_Age >= 18), reason = "Removed Attained_Age below 18")
ilec <- log_detailed_summary(ilec, quote(Attained_Age <= 90), reason = "Removed Attained_Age above 90")
ilec <- log_detailed_summary(ilec, quote(Issue_Age >= 18), reason = "Removed Issue_Age below 18")
ilec <- log_detailed_summary(ilec, quote(Issue_Age <= 80), reason = "Removed Issue_Age above 80")
ilec <- log_detailed_summary(ilec, quote(Duration <= 20), reason = "Removed duration above 20")
#ilec <- log_detailed_summary(ilec, quote(Insurance_Plan == "Term"), reason = "Term policies only") # Already taken out by pre-processing
ilec <- log_detailed_summary(ilec, quote(Smoker_Status != "U"), reason = "Removed unknown smoking status")

DT::datatable(attributes(ilec)$filter_summary)
```

## Remaining
- AtoEs based on SOA-supplied ExpDth_VBT2015wMI_Cnt that came with the dataset. <br>

```{r Data remaining}
summary <- 
  ilec[, list(
    Deaths = sum(Death_Count),
    Claim_Amounts_Bn = (sum(Death_Claim_Amount)/1000000000) %>% round(4),
    Exposure_years_Mln = (sum(Policies_Exposed)/1000000) %>% round(4),
    Avg_claim = (sum(Death_Claim_Amount)/sum(Death_Count)) %>% round(-3),
    AtoE_lives = (sum(Death_Count)/sum(ExpDth_VBT2015wMI_Cnt)) %>% round(4) * 100
  )]

DT::datatable(summary)
```


```{r aggregate to speed up analytics}
group_cols <- c("Observation_Year", "Sex", "Smoker_Status", "Issue_Age", "Duration", "Face_Amount_Band", "Issue_Year", "Attained_Age")
measure_cols <- c("Death_Count", "Policies_Exposed", "ExpDth_VBT2015wMI_Cnt")

ilec <- ilec[, lapply(.SD, sum, na.rm = TRUE), by = group_cols, .SDcols = measure_cols]
```

```{r covert chars to factors}
char_cols <- names(ilec)[vapply(ilec, is.character, logical(1))]
if (length(char_cols)) {
  ilec[, (char_cols) := lapply(.SD, as.factor), .SDcols = char_cols]
}

cols_to_order <- c("Face_Amount_Band")
for (j in intersect(cols_to_order, names(ilec))) {
  x  <- ilec[[j]]
  lv <- order_band_levels(x)
  f  <- factor(x, levels = lv, ordered = TRUE)
  set(ilec, j = j, value = f)
}
```

```{r test train split}
if(isTRUE(test_calendaryear)){
  years_sorted <- sort(unique(ilec$Observation_Year))
  test_years <- tail(years_sorted, 2)
  is_test <- ilec$Observation_Year %in% test_years
} else {
  ilec_rows <- nrow(ilec)
  test_rows <- sample(ilec_rows, floor(0.2 * ilec_rows))
  is_test <- rownames(ilec) %in% test_rows 
  }

dt_train <- ilec[!is_test, , drop = FALSE]
dt_test <- ilec[ is_test, , drop = FALSE]
```

```{r feature engineering}
age_coefs <- attr(poly(dt_train$Attained_Age, 3), "coefs")
entry_age_coefs <- attr(poly(dt_train$Issue_Age, 3), "coefs")
obs_year_coefs <- attr(poly(dt_train$Observation_Year, 3), "coefs")

dt_train <- gen_features(dt_train, age_coefs, entry_age_coefs, obs_year_coefs)
dt_test <- gen_features(dt_test, age_coefs, entry_age_coefs, obs_year_coefs)
```

```{r GLM baseline}
f <- list()
f$baseline <- stats::as.formula(
  Death_Count ~ (agepoly1 + agepoly2 + agepoly3 + agespline2 + agespline3 + agespline4 + Sex * Smoker_Status + Face_Amount_Band + issue_agepoly1 + issue_agepoly2 + Duration + invDur + dur_sq + dur_cube + obs_yearpoly1 + obs_yearpoly2)
  )
glm_baseline <- glm(as.formula(f$baseline), dt_train, family = "poisson", offset = log(Policies_Exposed))
```

# Baseline model
```{r }
DT::datatable(
  glm_rr_table(glm_baseline),
  rownames = FALSE,
  extensions = "Buttons",
  options = glm_dt_options()
) %>% glm_dt_format()
```

```{r add baseline model pred}
dt_train[, c("pred_baseline") := predict(glm_baseline, dt_train, newoffset = dt_train$Policies_Exposed) %>% exp]
dt_test[, c("pred_baseline") := predict(glm_baseline, dt_test, newoffset = dt_test$Policies_Exposed) %>% exp]
```

# Build XGBoost to probe the GLM fit for deficiencies {.tabset}
`r if(test_calendaryear) "- For the XGBoost, 2017 is used as test data, whilst 2012-2016 are train data <br>"`
- The baseline here (i.e. model being evaluated) is the GLM_baseline model <br>
-- This model was selected as a baseline to keep the workflow simple. 

## Model Training
```{r XGBoost, include = FALSE}
if(isTRUE(test_calendaryear)){
  years_sorted <- sort(unique(dt_train$Observation_Year))
  test_years <- tail(years_sorted, 1)
  is_validation <- dt_train$Observation_Year %in% test_years
} else {
  train_rows <- nrow(dt_train)
  test_rows <- sample(train_rows, floor(0.2 * train_rows))
  is_validation <- rownames(dt_train) %in% test_rows 
  }

feature_cols <- names(dt_train)[1:8]

dt_tr <- dt_train[!is_validation,]
dt_va <- dt_train[is_validation,]

# Targets and offsets
y_tr <- dt_tr[["Death_Count"]]
y_va <- dt_va[["Death_Count"]]
offset_tr <- log(dt_tr[["pred_baseline"]])
offset_va <- log(dt_va[["pred_baseline"]])

# Build a formula for the selected features, ex-intercept since we one-hot encode
f$XGBoost <- as.formula(
  paste("~", paste(feature_cols, collapse = " + "), "- 1"))

# One-hot encode
X_tr <- onehot_encode(dt_tr[, 1:8], f$XGBoost)
X_va <- onehot_encode(dt_va[, 1:8], f$XGBoost)

# DMatrices
dtrain <- xgb.DMatrix(data = X_tr, label = y_tr)
setinfo(dtrain, "base_margin", offset_tr)

dvalid <- xgb.DMatrix(data = X_va, label = y_va)
setinfo(dvalid, "base_margin", offset_va)

# Set interaction constraints within one-hot encoded features
onehot_cols <- colnames(X_tr)
face_amount_features <- grep("^Face_Amount_Band", onehot_cols, value = TRUE)
sex_features <- c("SexF", "SexM")
smoker_features <- c("Smoker_StatusNS", "Smoker_StatusS")
numeric_features <- c("Observation_Year", "Issue_Age", "Duration", "Issue_Year", "Attained_Age")

feature_groups <- list()
non_face_features <- c(sex_features, smoker_features, numeric_features)
feature_groups[[1]] <- which(onehot_cols %in% non_face_features) - 1

for(i in seq_along(face_amount_features)) {
  feature_groups[[i + 1]] <- which(onehot_cols == face_amount_features[i]) - 1
}
params <- list(
  objective = "count:poisson",
  eval_metric = "poisson-nloglik",
  eta = 0.1,
  max_depth = 3,
  subsample = 0.6,
  colsample_bytree = 0.8,
  min_child_weight = 500,
  interaction_constraints = feature_groups
  )

evals_list <- list(train = dtrain, validation = dvalid)

xgb_model <- xgb.train(
  params = params,
  data = dtrain,
  nrounds = 2000,
  evals = evals_list,
  early_stopping_rounds = 20,
  print_every_n = 10,
  verbose = 0
  )
```

## SHAP Feature Importance
- On training data <br>

```{r SHAP_importance}
shap_vals <- shap.values(xgb_model, X_tr)
exposure_weights <- exp(offset_tr)

shap_importance <- data.frame(
  Feature = colnames(shap_vals$shap_score),
  MeanAbsSHAP = apply(shap_vals$shap_score, 2, function(x) mean(abs(x))) %>% round(6),
  MeanAbsWeightedSHAP = apply(shap_vals$shap_score, 2, function(x) weighted.mean(abs(x), exposure_weights)) %>% round(6)
) %>% data.table

shap_importance <- shap_importance[order(-shap_importance$MeanAbsWeightedSHAP), ]
DT::datatable(shap_importance)
```

## SHAP Beeswarm Plot
- On training data <br>

```{r shap_summary, fig.width=10, fig.height=8}
shap_long <- SHAPforxgboost::shap.prep(
  shap_contrib = shap_vals$shap_score,
  X_train = X_tr
)

SHAPforxgboost::shap.plot.summary(shap_long)
```

## SHAP Dependence Plots{.tabset}
- On training data. <br>
- SHAP is computed on a weighted random sample of 2,000 rows, where sampling probabilities are proportional to life‑year exposure. <br>

```{r SHAP dependence, results='asis', echo = F, message = F, warning = F}
sub_idx <- sample(unique(shap_long$ID), size = 2000, prob = dt_tr$Policies_Exposed)
N_features <- pmin(nrow(shap_importance), 10)

for(feature in shap_importance$Feature[1:N_features]){
  cat(paste0("### ", feature," \n\n"))
  
  plot <- shap.plot.dependence(
    shap_long[ID %in% sub_idx],
    X = X_tr[sub_idx,],
    x = feature,
    color_feature = feature_cols[2]
  )
  
#  plot <- plot + 
#    scale_y_log10(
#      labels = function(x) round(exp(x), 4),
#      name = "exp(SHAP value), A/E equivalent"
#    )
  feature_values <- X_tr[sub_idx, feature, with = FALSE] %>% unlist
  if(length(unique(feature_values)) == 2 && all(feature_values %in% c(0, 1, TRUE, FALSE))) {
    plot <- plot + 
      scale_x_continuous(
        breaks = c(0, 1),
        labels = c("FALSE", "TRUE")
      )
  }
  
  plotly_obj <- ggplotly(plot)
  print(htmltools::tagList(plotly_obj))
  cat('\n\n')
}
```

## SHAP Interaction Analysis
```{r shap_interactions, fig.width=12, fig.height=10}
shap_int <- predict(xgb_model, X_tr, predinteraction = TRUE)

interaction_summary <- data.frame()
exposure_weights <- exp(offset_tr)

for(i in 1:(ncol(X_tr)-1)) {
  for(j in (i+1):ncol(X_tr)) {
    feature1 <- colnames(X_tr)[i]
    feature2 <- colnames(X_tr)[j]
    interaction_values <- shap_int[, i, j]
    
    interaction_summary <- rbind(
      interaction_summary,
      data.table(
        Feature1 = feature1,
        Feature2 = feature2,
        MeanAbsInteraction = mean(abs(interaction_values)) %>% round(6),
        WeightedMeanAbsInteraction = weighted.mean(abs(interaction_values), exposure_weights) %>% round(6)
      )
    )
  }
}

interaction_summary <- interaction_summary[MeanAbsInteraction != 0]
interaction_summary <- interaction_summary[order(-interaction_summary$WeightedMeanAbsInteraction), ]
DT::datatable(interaction_summary, options = list(pageLength = 20))
```

## SHAP Interaction Dependence Plots{.tabset}
- On training data. <br>
- SHAP is computed on a weighted random sample of 2,000 rows, where sampling probabilities are proportional to life‑year exposure. <br>

```{r SHAP_interaction_dependence, results='asis', echo = F, message = F, warning = F}
sub_idx <- sample(unique(shap_long$ID), size = 2000, prob = dt_tr$Policies_Exposed)

N_interactions <- pmin(nrow(interaction_summary), 10)
top_interactions <- interaction_summary[1:N_interactions]

for(k in 1:nrow(top_interactions)) {
  feature1 <- top_interactions$Feature1[k]
  feature2 <- top_interactions$Feature2[k]
  
  cat(paste0("### ", feature1, " × ", feature2, " \n\n"))
  
  idx1 <- which(colnames(X_tr) == feature1)
  idx2 <- which(colnames(X_tr) == feature2)
  
  interaction_vals <- shap_int[sub_idx, idx1, idx2]
  
  plot_data <- data.frame(
    Feature1_Value = X_tr[sub_idx][[feature1]],
    Feature2_Value = X_tr[sub_idx][[feature2]],
    Interaction_SHAP = interaction_vals
  )
  
  plot <- ggplot(plot_data, aes(x = Feature1_Value, y = Feature2_Value, color = Interaction_SHAP)) +
    geom_point(alpha = 0.6, size = 1.5) +
    scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
    labs(
      x = feature1,
      y = feature2,
      color = "Interaction\nSHAP Value",
      title = paste("SHAP Interaction:", feature1, "×", feature2)
    ) +
    theme_minimal()
  
  feature1_values <- X_tr[sub_idx][[feature1]]
  feature2_values <- X_tr[sub_idx][[feature2]]
  
  if(length(unique(feature1_values)) == 2 && all(feature1_values %in% c(0, 1, TRUE, FALSE))) {
    plot <- plot + 
      scale_x_continuous(breaks = c(0, 1), labels = c("FALSE", "TRUE"))
  }
  
  if(length(unique(feature2_values)) == 2 && all(feature2_values %in% c(0, 1, TRUE, FALSE))) {
    plot <- plot + 
      scale_y_continuous(breaks = c(0, 1), labels = c("FALSE", "TRUE"))
  }
  
  plotly_obj <- ggplotly(plot)
  print(htmltools::tagList(plotly_obj))
  cat('\n\n')
}
```

## Interaction Heatmap
```{r SHAP dependence interactions, results='asis', echo = F, message = F, warning = F}
# Create interaction matrix (approximation)
#if (length(top_features_interact) >= 2) {
#  # Calculate pairwise SHAP interaction strengths
#  interaction_matrix <- matrix(0, 
 #                               nrow = length(top_features_interact), 
#                                ncol = length(top_features_interact))
#  rownames(interaction_matrix) <- top_features_interact
#  colnames(interaction_matrix) <- top_features_interact
  
#  for (i in 1:length(top_features_interact)) {
#    for (j in 1:length(top_features_interact)) {
#      if (i != j) {
#        # Correlation of SHAP values as proxy for interaction
#        interaction_matrix[i, j] <- abs(cor(
#          shap_values$shap_score[, top_features_interact[i]],
#          shap_values$shap_score[, top_features_interact[j]]
#        ))
#      }
#    }
#  }
  
  # Plot heatmap
#  library(reshape2)
#  melted_matrix <- melt(interaction_matrix)
  
#  ggplot(melted_matrix, aes(x = Var1, y = Var2, fill = value)) +
#    geom_tile() +
#    geom_text(aes(label = round(value, 2)), size = 3) +
#    scale_fill_gradient2(low = "white", high = "red", mid = "yellow", 
#                         midpoint = 0.5, limit = c(0, 1)) +
#    theme_minimal() +
#    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
#    labs(
#      title = "SHAP Value Correlation Matrix (Interaction Proxy)",
#      subtitle = "Identifying potential interaction terms missing from GLM_basic",
#      x = "", y = "", fill = "Abs Correlation"
#    )
#}
```

```{r add xgboost pred, include = FALSE}
y_train_full <- dt_train[["Death_Count"]]
y_test <- dt_test[["Death_Count"]]
offset_tr <- log(dt_train[["pred_baseline"]])
offset_test <- log(dt_test[["pred_baseline"]])

# One-hot encode
X_train_full<- onehot_encode(dt_train[, 1:8], f$XGBoost)
X_test <- onehot_encode(dt_test[, 1:8], f$XGBoost)

# DMatrices
dtrain_full <- xgb.DMatrix(data = X_train_full, label = y_train_full)
setinfo(dtrain_full, "base_margin", offset_tr)

dtest <- xgb.DMatrix(data = X_test, label = y_test)
setinfo(dtest, "base_margin", offset_test)

dt_train[, c("pred_xgboost") := predict(xgb_model, dtrain_full)]
dt_test[, c("pred_xgboost") := predict(xgb_model, dtest)]
```

# GLM supplemented with interactions identified by XGBoost
```{r GLM interactions}
interaction_terms <- top_interactions[, {
  feat1 <- sapply(feature_cols, function(x) grepl(paste0("^", x), Feature1))
  feat2 <- sapply(feature_cols, function(x) grepl(paste0("^", x), Feature2))
  
  feature1_clean <- feature_cols[apply(feat1, 1, which.max)]
  feature2_clean <- feature_cols[apply(feat2, 1, which.max)]
  
   replacements <- list(
        Issue_Age = "(issue_agepoly1 + issue_agepoly2)",
        Attained_Age = "(agepoly1 + agepoly2)",
        Duration = "(Duration + invDur + dur_sq)"
    )
   
    feature1_expanded <- ifelse(feature1_clean %in% names(replacements),
                                replacements[feature1_clean],
                                feature1_clean)
    feature2_expanded <- ifelse(feature2_clean %in% names(replacements),
                                replacements[feature2_clean],
                                feature2_clean)
    
    paste(feature1_expanded, feature2_expanded, sep = ":")
    }] %>% unique

baseline_rhs <- as.character(f$baseline)[3]
baseline_rhs_clean <- gsub("^\\((.*)\\)$", "\\1", baseline_rhs)
all_terms <- paste(c(baseline_rhs_clean, interaction_terms), collapse = " + ")
formula_string <- paste("Death_Count ~", all_terms)
f$baseline_interactions <- stats::as.formula(formula_string)

glm_baseline_interactions <- glm(f$baseline_interactions, dt_train, 
                                  family = "poisson", offset = log(Policies_Exposed))
```

## Baseline model
```{r }
DT::datatable(
  glm_rr_table(glm_baseline_interactions),
  rownames = FALSE,
  extensions = "Buttons",
  options = glm_dt_options()
) %>% glm_dt_format()
```

```{r add baseline interactions model pred}
dt_train[, c("pred_baseline_interactions") := predict(glm_baseline_interactions, dt_train, newoffset = dt_train$Policies_Exposed) %>% exp]
dt_test[, c("pred_baseline_interactions") := predict(glm_baseline_interactions, dt_test, newoffset = dt_test$Policies_Exposed) %>% exp]
```


## Model Comparison
```{r }

```

# LASSO supplemented with interactions identified by XGBoost
```{r LASSO}
X_train <- model.matrix(f$baseline_interactions, dt_train)[, -1]
y_train <- dt_train$Death_Count
offset_train <- log(dt_train$Policies_Exposed)

cv_baseline_interactions <- cv.glmnet(
    x = X_train,
    y = y_train,
    family = "poisson",
    offset = offset_train,
    alpha = 1,  # 1 = lasso, 0 = ridge, 0 < alpha < 1 = elastic net
    nfolds = 10
)

dt_train[, c("pred_LASSO_baseline_interactions") := 
           predict(cv_baseline_interactions, 
                      newx = X_train, 
                      newoffset = offset_train,
                      s = "lambda.min",
                      type = "response")]

X_test <- model.matrix(f$baseline_interactions, dt_test)[, -1]
offset_test <- log(dt_test$Policies_Exposed)

dt_test[, c("pred_LASSO_baseline_interactions") := 
           predict(cv_baseline_interactions, 
                      newx = X_test, 
                      newoffset = offset_test,
                      s = "lambda.min", 
                      type = "response")]
```

# IBLM models
- IBLM models were build with tree depth 1, 2 and 3 on the booster XGBoost model. <br>
-- This decision was made as it was felt that a tree depth of 3 (the default), results in (very) hard to explain third order variable interactions. 

```{r IBLM}
xcols <- c(names(dt_tr[, 1:9]), "Policies_Exposed")

base_params <- list(
  seed = 1234,
  eta = 0.1,
  subsample = 0.6,
  colsample_bytree = 0.8,
  min_child_weight = 500
)

df_list <- list(
    train = dt_tr[, ..xcols], # Same as used for XGBoost
    test = dt_test[, ..xcols],
    validate = dt_va[, ..xcols] # Same as used for XGBoost
    )

iblm_models <- setNames(vector("list", length(depths)), paste0("d", depths))
for (d in depths) {
  iblm_models[[paste0("d", d)]] <- train_iblm_xgb_offset(
    df_list,
    response_var = "Death_Count",
    family = "poisson",
    params = c(base_params, list(max_depth = d)),
    offset_var = "Policies_Exposed"
  )
}

for (d in depths) {
  nm <- paste0("d", d)
  pred_col <- paste0("pred_IBLM_", nm)
  
  dt_train[, (pred_col) := predict.iblm(iblm_models[[nm]],
                                        dt_train[, ..xcols])]
  
  dt_test[,  (pred_col) := predict.iblm(iblm_models[[nm]],
                                        df_list$test)]
}
```

```{r Save data and splits for collaborators}
#ilec$is_test <- is_test
#ilec$is_validation <- FALSE
#ilec[is_test == FALSE]$is_validation <- is_validation
#arrow::write_parquet(ilec, "Data/ILEC_2012_19_aggregated_20260123.parquet")
```

# Visuals of model fits {.tabset}
```{r, results = 'asis', echo = FALSE}
models <- c("Baseline GLM" = "pred_baseline", 
            "XGBoost" = "pred_xgboost", 
            "IBLM_depth2" = "pred_IBLM_d2", 
            "ILEC_Basis" = "ExpDth_VBT2015wMI_Cnt")

actual_col <- "Death_Count"

for(model_name in names(models)) {
  cat(paste0("\n## ", model_name, " {.tabset}\n\n"))
  
  if(test_calendaryear) {
    cat("- Visuals on the train data (calendar years 2012-2017)\n")
    cat("- Visuals on the test data (calendar years 2018-2019)\n\n")
  }
  
  pred_col <<- models[model_name]
  
  ae_plots <- setNames(
    lapply(group_cols, function(gc) {
      summ <- ae_summary_by(dt_train, dt_test, gc)
      ae_plot(summ, gc, top_n = 50)
    }),
    group_cols
  )
  
  for(i in seq_along(ae_plots)) {
    cat(paste0("\n### ", names(ae_plots)[i], "\n\n"))
    print(htmltools::tagList(plotly::ggplotly(ae_plots[[i]])))
    cat("\n")
  }
}
```

# Evaluate model performance
- Poisson deviance is the correct metric, however standardisation is required in order to compare train and test data performance. <br>
-- Average Poisson deviance is incorrect since exposure per row ranges from a single day to 5,502 years. <br> <br>

<b>Metrics</b>
-- <b>Preferred Metric</b>: mean Poisson deviance, weighted by ILEC expected claims (i.e. Poisson deviance per expected claim). <br>
-- Metric: mean Poisson deviance, weighted by life year exposure (i.e. Poisson deviance per predicted claim). <br>
--- Downside is that low risk life years (e.g. 20 year olds) given the same weight as high risk life years (e.g. 80 year olds) <br>
-- Metric: mean Poisson deviance, weighted by actual claims (i.e. Poisson deviance per predicted claim). <br>
--- Downside is that entries without claims are ignored, so this metric does not protect well against overprediction of claims. <br><br>

``` {r, results = 'asis', echo = F}
dev_term <- stats::poisson()$dev.resids

pred_spec <- c(
  GLM_baseline                 = "pred_baseline",
  GLM_baseline_interactions    = "pred_baseline_interactions",
  LASSO_baseline_interactions  = "pred_LASSO_baseline_interactions",
  XGBoost                      = "pred_xgboost",
  ILEC_basis                   = "ExpDth_VBT2015wMI_Cnt"
)

iblm_spec <- setNames(
  paste0("pred_IBLM_d", depths),
  paste0("IBLM_d", depths)
)

pred_spec <- c(pred_spec, iblm_spec)

models <- imap(pred_spec, ~ list(
  train_pred = dt_train[[.x]],
  test_pred  = dt_test[[.x]]
))

weights <- list(
  unweighted = list(train = rep(1, nrow(dt_train)), test = rep(1, nrow(dt_test))),
  wpred      = list(train = dt_train$ExpDth_VBT2015wMI_Cnt, test = dt_test$ExpDth_VBT2015wMI_Cnt),
  wexposure  = list(train = dt_train$Policies_Exposed,      test = dt_test$Policies_Exposed),
  wclaims    = list(train = dt_train$Death_Count,           test = dt_test$Death_Count)
)

mean_pois_dev <- function(y, mu, wt) mean(dev_term(y, mu, wt = wt))

comparison <- imap_dfr(models, function(m, nm) {
  data.table(
    Model               = nm,
    avgPoisdev_train    = mean_pois_dev(dt_train$Death_Count, m$train_pred, weights$unweighted$train),
    avgPoisdev_test     = mean_pois_dev(dt_test$Death_Count,  m$test_pred,  weights$unweighted$test),
    Poisdev_train_wpred = mean_pois_dev(dt_train$Death_Count, m$train_pred, weights$wpred$train),
    Poisdev_test_wpred  = mean_pois_dev(dt_test$Death_Count,  m$test_pred,  weights$wpred$test),
    Poisdev_train_wexposure = mean_pois_dev(dt_train$Death_Count, m$train_pred, weights$wexposure$train),
    Poisdev_test_wexposure  = mean_pois_dev(dt_test$Death_Count,  m$test_pred,  weights$wexposure$test),
    Poisdev_train_wclaims   = mean_pois_dev(dt_train$Death_Count, m$train_pred, weights$wclaims$train),
    Poisdev_test_wclaims    = mean_pois_dev(dt_test$Death_Count,  m$test_pred,  weights$wclaims$test)
  )
})

DT::datatable(comparison) |>
  DT::formatSignif(
    columns = setdiff(names(comparison), "Model"),
    digits = 4
  )
```

