---
title: "ILEC data analysis"
author: "Michiel Luteijn"
date: '2025-12-16'
output:
  html_document:
    toc: yes
    toc_float: yes
    code_folding: hide
---

# Background  {.tabset}

## Objectives
**Big Pitch submission** <br>
Health and care (H&C) actuaries are well positioned to benefit from recent advances in data science, as machine learning (ML) techniques have become increasingly transparent and accessible. <br>

Once seen as opaque, explainable ML now enables actuaries to improve predictive power whilst maintaining interpretability. Explainable ML can enhance traditional methods, whilst retaining the ability to infuse actuarial judgement.  <br>

We will demonstrate that H&C actuaries can have the best of both worlds by combining transparency, expert knowledge, and modern analytical innovation.

**IFoA webpage** <br>
TITLE: Breaking down the black box: why actuaries should embrace machine learning now. <br>
Description: Achieving the best of both worlds with explainable machine learning [H&C: Data Techniques in H&C WP].

## Decisions made
**Study design** <br>
- It was decided in Dec to build incidence models (i.e. from scratch with lives exposure as the offset) <br>
- Dec 19th - Provisionally decided to have the last 2 calendar years (2018-2019) as the holdout

**Study inclusion criteria** <br>
- Dec 19th - decided TERM only <br>
- Dec 19th - to take out fields with no exposure (none of these have claims) <br>
- Dec 19th - decided the cut-off for issue year should be 1980 onwards <br>
- Dec 19th - decided the all observation years (2012-2019) will be included <br>
- Dec 19th - decided the range for issueAge should be 18-80 <br>
- Dec 19th - decided the range for Attained_Age should be 18-90 <br>
- Dec 19th - No exclusions made based on combinations of SOA_Guar_Lvl_TP and duration (for now), or Face_Amount_Band <br>
- Jan 9th - decided the cap duration at 20 due to concerns about data quality at (very) high durations and low volumes of data
- Jan 13th - decided the exclude unknown smoking status


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r package install, results=FALSE, message=FALSE, warning=FALSE, include = F}
packages <- c("data.table", "dplyr", "ggplot2", "magrittr", "purrr", "knitr", "stringr", "plyr", "htmlTable", "lme4", "statmod", "glmnet", "splines", "MASS", "rms", "caret", "foreach", "parallel", "readr", "arrow", "doSNOW", "here", "htmltools", "plotly")

for (pkg in packages){
  if(!pkg %in% rownames(installed.packages())){
    install.packages(pkg)
  } 
}
library("purrr")
quietly(lapply(packages, library, character.only = T)) %>% suppressWarnings()

rm(packages, pkg)
quietly(gc())
```

```{r, include=FALSE, echo=FALSE}
library(plotly)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(fig.showtext = TRUE)
```

```{r, include=FALSE, echo=FALSE}
# Source https://stackoverflow.com/questions/60685631/using-ggplotly-and-dt-from-a-for-loop-in-rmarkdown
# Init Step to make sure that the dependencies are loaded
htmltools::tagList(DT::datatable(cars))
htmltools::tagList(ggplotly(ggplot()))
```

```{r Register parallel backend, include=FALSE}
if(!exists("cl")){
  cl <- parallel::makeCluster(8, type = "SOCK") # Creating a SOCK cluster
  doSNOW::registerDoSNOW(cl)
}
```

```{r functions, include=FALSE}
helpers <- list.files(path = here::here("R/Helpers"), pattern = "\\.R$", full.names = TRUE)
lapply(helpers, source)
```

```{r Data load, include=FALSE}
fpath = "Data/ILEC_2012_19 - 20240429.parquet"

if(!file.exists(here::here(fpath))){
  source("R/read_data_and_save_parquet.R")
  } else {
    ilec <- arrow::read_parquet(here::here(fpath))
    }
```

# Data filtering
- AtoEs based on SOA-supplied ExpDth_VBT2015wMI_Cnt that came with the dataset.
- For age nearest birthday, half a year is subtracted from AttainedAge prior to AttainedAge filters

```{r Data filters}
ilec <- log_detailed_summary(ilec, quote(!Policies_Exposed == 0), reason = "Taking out zero exposure fields")
ilec <- log_detailed_summary(ilec, quote(Issue_Year >= 1980), reason = "Removed IssueYear pre-1980")
ilec <- log_detailed_summary(ilec, quote(Attained_Age >= 18), reason = "Removed Attained_Age below 18")
ilec <- log_detailed_summary(ilec, quote(Attained_Age <= 90), reason = "Removed Attained_Age above 90")
ilec <- log_detailed_summary(ilec, quote(Issue_Age >= 18), reason = "Removed Issue_Age below 18")
ilec <- log_detailed_summary(ilec, quote(Issue_Age <= 80), reason = "Removed Issue_Age above 80")
ilec <- log_detailed_summary(ilec, quote(Duration <= 20), reason = "Removed duration above 20")
ilec <- log_detailed_summary(ilec, quote(Smoker_Status != "U"), reason = "Removed unknown smoking status")

DT::datatable(attributes(ilec)$filter_summary)
```

```{r aggregate to speed up analytics}
group_cols <- c("Observation_Year", "Sex", "Smoker_Status", "Issue_Age", "Duration", "Face_Amount_Band", "Issue_Year", "Attained_Age")
measure_cols <- c("Death_Count", "Policies_Exposed", "ExpDth_VBT2015wMI_Cnt")

ilec <- ilec[, lapply(.SD, sum, na.rm = TRUE), by = group_cols, .SDcols = measure_cols]
```

```{r covert chars to factors}
char_cols <- names(ilec)[vapply(ilec, is.character, logical(1))]
if (length(char_cols)) {
  ilec[, (char_cols) := lapply(.SD, as.factor), .SDcols = char_cols]
}

cols_to_order <- c("Face_Amount_Band")
for (j in intersect(cols_to_order, names(ilec))) {
  x  <- ilec[[j]]
  lv <- order_band_levels(x)
  f  <- factor(x, levels = lv, ordered = TRUE)
  set(ilec, j = j, value = f)
}
```

```{r test train split}
years_sorted <- sort(unique(ilec$Observation_Year))
test_years <- tail(years_sorted, 2)
is_test <- ilec$Observation_Year %in% test_years
dt_train <- ilec[!is_test, , drop = FALSE]
dt_test <- ilec[ is_test, , drop = FALSE]
```

```{r feature engineering}
age_coefs <- attr(poly(dt_train$Attained_Age, 3), "coefs")
entry_age_coefs <- attr(poly(dt_train$Issue_Age, 3), "coefs")
obs_year_coefs <- attr(poly(dt_train$Observation_Year, 3), "coefs")

dt_train <- gen_features(dt_train, age_coefs, entry_age_coefs, obs_year_coefs)
dt_test <- gen_features(dt_test, age_coefs, entry_age_coefs, obs_year_coefs)
```

# Basic GLM {.tabset}
- First build age only. The age only model will be the offset for models downstream to allow for better interpretation of regression coefficients. 

```{r GLM}
f <- list()
f$age_fit <- stats::as.formula(
  Death_Count ~ 
    (agepoly1 + agepoly2 + agepoly3 + agespline2 + agespline3 + agespline4)
)
glm_age_fit <- glm(as.formula(f$age_fit), dt_train, family = "poisson", offset = log(Policies_Exposed))
```

## Age model
```{r }
DT::datatable(
  glm_rr_table(glm_age_fit),
  rownames = FALSE,
  extensions = "Buttons",
  options = glm_dt_options()
) %>% glm_dt_format()
```

```{r add pred}
dt_train[, c("age_pred") := predict(glm_age_fit, dt_train, newoffset = dt_train$Policies_Exposed) %>% exp]
dt_test[, c("age_pred") := predict(glm_age_fit, dt_test, newoffset = dt_test$Policies_Exposed) %>% exp]
```

```{r GLM basic}
f$basic <- stats::as.formula(
  Death_Count ~ (Sex * Smoker_Status + Face_Amount_Band + issue_agepoly1 + issue_agepoly2 + Duration + invDur + dur_sq + dur_cube + obs_yearpoly1 + obs_yearpoly2)
  )
glm_basic <- glm(as.formula(f$basic), dt_train, family = "poisson", offset = log(age_pred))
```

## Stacked basic model
```{r }
DT::datatable(
  glm_rr_table(glm_basic),
  rownames = FALSE,
  extensions = "Buttons",
  options = glm_dt_options()
) %>% glm_dt_format()
```

```{r add age model pred}
dt_train[, c("basic_pred") := predict(glm_basic, dt_train, newoffset = dt_train$age_pred) %>% exp]
dt_test[, c("basic_pred") := predict(glm_basic, dt_test, newoffset = dt_test$age_pred) %>% exp]
```

# Visuals of model fit {.tabset}
- Visuals on the test data (calendar years 2018-2019)

## Age model {.tabset}
- Note AtoEs are low since experience is lighter in 2018-2019, then prior observation years which is not accounted for in the age model <br>.

```{r }
ae_plots  <- setNames(vector("list", length(group_cols)), group_cols)
actual_col <- "Death_Count"
pred_col   <- "age_pred"

ae_plots <- setNames(
  lapply(group_cols, function(gc) {
    summ <- ae_summary_by(dt_test, gc)
    ae_plot(summ, gc, top_n = 50)
  }),
  group_cols
)
```

``` {r, results = 'asis', echo = F}
for(i in 1:length(ae_plots)){
  cat(paste0("### ", names(ae_plots)[i]," \n"))
  print(htmltools::tagList(plotly::ggplotly(ae_plots[[i]])))
}
```

## Basic GLM {.tabset}

```{r }
ae_plots  <- setNames(vector("list", length(group_cols)), group_cols)
pred_col   <- "basic_pred"

ae_plots <- setNames(
  lapply(group_cols, function(gc) {
    summ <- ae_summary_by(dt_test, gc)
    ae_plot(summ, gc, top_n = 50)
  }),
  group_cols
)
```

``` {r, results = 'asis', echo = F}
for(i in 1:length(ae_plots)){
  cat(paste0("### ", names(ae_plots)[i]," \n"))
  print(htmltools::tagList(plotly::ggplotly(ae_plots[[i]])))
}
```

