---
title: IFoA DS in Health and Care Case Study on CMI Data
author: "Michiel Luteijn"
date: <span style="color:#005192">`r Sys.Date()`</span><br><br/>
output: 
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: 3
    number_sections: true
    code_folding: hide
    df_print: paged
    fig_caption: yes
---

<style>
  .plotly {
    margin-bottom: 400px;
  }
</style>

```{r package install, results=FALSE, message=FALSE, warning=FALSE, include = F}
packages <- c("data.table", "dplyr", "ggplot2", "magrittr", "purrr", "knitr", "plyr", "here", "plotly", "htmltools", "DT", "ggcorrplot", "Hmisc")

for (pkg in packages){
  if(!pkg %in% rownames(installed.packages())){
    install.packages(pkg)
  } 
}
library("purrr")
quietly(lapply(packages, library, character.only = T)) %>% suppressWarnings()

rm(packages, pkg)
quietly(gc())
```

```{r general parameters, echo = FALSE}
# Some general settings for the .html.

knitr::opts_chunk$set(
  echo = FALSE, # code is not shown in document 
  warning = FALSE, # excludes warning messages to be displayed
  message = FALSE, # excludes messages to be displayed
  # set figure size and position
  fig.align = 'center',
  fig.asp = 0.618,
  fig.width = 8.6,
  out.width='100%'
)

```

# Load in data
```{r load CMI data}
homedir <- here::here()
set.seed(123)

# Directory for CMI data
datadir <- paste0(homedir,"/Imports/Data/")
# Directory for base rate tables
tabledir <- paste0(homedir,"/Imports/Tables/")
# Directory for trends
trenddir <- paste0(homedir,"/Imports/Trend/")

# Read in the CMI data for 2007-2019 - objective = expdata
source(paste0(homedir, "/R/Helpers/Read CMI data.R"))
```

```{r format data}
expdata[, ':=' (
  commencementyear = as.integer(commencementyear),
  duration = as.integer(duration),
  year = as.integer(year),
  sumassuredband = factor(sumassuredband, levels = c("£0-£25000", "£25001-£75000", "£75001-£125000", "£125001-£250000", "£250001+"), ordered = TRUE)
  )]
```

# Data exploration


## Correlation matrix {.tabset}
- Weighted by lives exposure
```{r predictors}
predictors <- names(expdata[, c(1:9, 18)])
```

### Plot
- It was observed that <br>
-- age is negatively correlated with sum assured band <br>
-- Single lives are positively correlated with level policies and commencement year <br>
-- Smoker status S is negatively correlated with sum assured band <br>
-- Distribution channel S is negatively correlated with commencement year and positively correlated with duration  <br>
-- Distribution channel U is positively correlated with product type Other and commencement year

```{r , echo=FALSE, results = 'asis'}
cormatdata <- data.table::copy(expdata)
cormatdata[, c("sumassuredband") := as.numeric(sumassuredband)]

formula <- as.formula(paste("incurredclaims ~", paste(predictors, collapse = " + ")))
model_matrix <- model.matrix(formula, cormatdata)
cor_matrix <- rcorr(model_matrix)

corr <- stats::cov.wt(model_matrix, 
                      wt = cormatdata$livesexposure, cor = TRUE) %>% .[["cor"]]

corr[upper.tri(corr)] <- NA  # Remove upper triangle
corr <- as.data.frame(as.table(corr)) %>% data.table
corr <- corr[Var1 != Var2 & stats::complete.cases(Freq) & Var1 != "(Intercept)" & Var2 != "(Intercept)"]

# Extract comparisons within one-hot encoded field
extract_base_column <- function(text_string, base_columns = names(expdata)) {
  base_columns[sapply(base_columns, function(base) grepl(base, text_string))]
  }
field1_base_columns <- sapply(corr$Var1, extract_base_column)
field2_base_columns <- sapply(corr$Var2, extract_base_column)

base_comparison <- sapply(1:length(field1_base_columns), function(i) {
  identical(field1_base_columns[[i]], field2_base_columns[[i]])
})

cormatrix <- ggplot(corr[!base_comparison], aes(Var1, Var2, fill = Freq)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white",
                       midpoint = 0, limit = c(-1, 1), space = "Lab",
                       name="Correlation") +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, vjust = 1, hjust = 1), axis.title.x = ggplot2::element_blank(),
                 axis.title.y = ggplot2::element_blank(), panel.grid.major = ggplot2::element_blank(), legend.justification = c(1, 0),
                 legend.position.inside = c(0.6, 0.7)) +
  ggplot2::geom_text(ggplot2::aes(Var1, Var2, label = round(Freq, 2)), color = 'black',size = 3) +
  coord_fixed() +
  ggtitle("Correlation Matrix")

plotlyOut <- plotly::ggplotly(cormatrix)
plotlyOut <- plotlyOut %>% layout(width = 1200, height = 900)
plotlyOut
```

### Table
```{r , echo=FALSE, results = 'asis'}
names(corr) <- c("Field1", "Field2", "Correlation")
corr <- corr[order(-abs(Correlation))]
corr[, c("Correlation") := round(Correlation, 2)]

DT::datatable(corr[!base_comparison])
```

## Univariate A/E plots {.tabset}

```{r , echo=FALSE, results = 'asis'}
for (i in 1:length(predictors)){
  l <- htmltools::tagList()
  l[[1]] <- list(cat(paste0('### ', predictors[i]), ' \n'))
  plotData <- expdata[, list(
    actual_L = sum(incurredclaims),
    expected_L = sum(expectedclaims),
    actual_A = sum(amountincurred),
    expected_A = sum(expectedamountclaims),
    exposure_L = sum(livesexposure),
    exposure_A = sum(amountsexposure)
  ), by = eval(predictors[i])]
  classPredictor <- class(expdata[, get(predictors[i])])[1]
  if(classPredictor %in% c("numeric", "integer", "Date")){
    plotOut <- ggplot2::ggplot() +
      ggplot2::geom_line(data = plotData,
                         mapping = ggplot2::aes(x = get(predictors[i]), y = actual_L/expected_L), lwd = 2) 
    
  } else{
    plotOut <- ggplot2::ggplot() +
      ggplot2::geom_point(data = plotData,
                          mapping = ggplot2::aes(x = get(predictors[i]), y = actual_L/expected_L))
  }
  plotOut <- plotOut +
    ggplot2::xlab(paste0(predictors[i])) +
    ggplot2::scale_y_continuous(limits = c(min(plotData$actual_L / plotData$expected_L, 1), max(plotData$actual_L / plotData$expected_L, 1))) +
    ggplot2::geom_hline(yintercept = 1, linetype = "dashed", color = "red")
  plotlyOut <- plotly::ggplotly(plotOut)
  l[[2]] <- list(plotlyOut)
  l[[3]] <- list(DT::datatable(plotData))
  print(l)
}

```







